---
title: "Creating a Draft Pick Value Chart"
author: "Tirdod Behbehani"
date: July 5, 2021
output: rmarkdown::github_document
always_allow_html: true
tables: true
---

## Project Description: Quantifying the Value of Each Pick in the NBA Draft

Looking at the drafts from 2001 through 2016, I used the Value Over 
Replacement Player (VORP) metric to find the mean impact for each pick in the draft.  Unlike performance metrics such as RAPM and PIPM, VORP reflects that players who don't play at all aren't helping their teams.  By using fifteen years of draft data, this provides 831 observations, which provides me with enough of a sample size for each pick to perform my analysis.

To grade each player, I took the sum of each player's VORP for each season they played under their rookie contract (four years for first round picks and three years for second round picks).  Using the sum, as opposed to the average, puts greater weight on players that played stayed in the league for more seasons than those who didn't pan out.

To account for random variation, I smoothed the pick values.  Additionally,
the pick values were normalized, setting the value of the first pick to 100.

## Load Modules, Import Data, and Clean Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='Figs/')
```

```{r echo=TRUE, results='hide'}
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(knitr)))
suppressWarnings(suppressMessages(library(kableExtra)))
suppressWarnings(suppressMessages(library(scam)))

draft <- suppressMessages(read_csv('draftHistory.csv'))
vorp <- suppressMessages(read_csv('vorp.csv'))
```
Drop Unnecessary Columns in the Draft data frame.

```{r echo=TRUE}
drops <- c('numberRoundPick', 'typeOrganizationFrom', 
           'cityTeam', 'teamName', 'slugOrganizationTypeFrom', 
           'locationOrganizationFrom', 'nameTeam', 'slugTeam', 'typeOrganizationFrom',
           'idTeam', 'nameOrganizationFrom')

draft <- select(draft, -c(all_of(drops)))
```
Let's show a brief snapshot of the two dataframes.

```{r echo=TRUE}
head(draft)
head(vorp)
```
Let's make a change to the VORP data frame.  Let's change any negative VORP value to zero.  As explained to me by Jake Flancer, when using replacement-player metrics, a pick can't have negative value since the player could be cut for a replacement level player instead.  

```{r echo=TRUE}
vorp$VORP <- ifelse(vorp$VORP < 0, 0, vorp$VORP)
```
In examining the two datasets, I noticed that there was one discrepancy between the two of them.  The VORP dataset used apostrophes in names (ex. De'Andre Hunter), whereas the draft dataset did not (ex. DeAndre Hunter).  With the sapply function, I made sure that the names in the two datasets matched one another.

```{r echo=TRUE}
vorp$Player <-  sapply(vorp$Player, function(x) gsub("'", "", x))
```

```{r include=FALSE}
vorp$Player <-  sapply(vorp$Player, function(x) gsub("Kristaps Porziņģis", "Kristaps Porzingis", x))
vorp$Player <-  sapply(vorp$Player, function(x) gsub("Skal Labissière", "Skal Labissiere", x))

```
We're only interested in a player's first four seasons (at most), so we can 
delete all seasons past a player's fourth.

```{r echo=TRUE}
vorp$count <- with(vorp, ave(vorp$Player, vorp$Player, FUN = seq_along))

vorp$count <- as.numeric(vorp$count)

stats <- subset(vorp, vorp$count <= 4)
```

Now, let's join the VORP data with the draft history.  We'll temorarily break
the data into first and second round picks so we can remove the fourth season
for second round picks.  Then we'll put it back into one dataframe.

```{r echo=TRUE}
firstRound <- subset(draft, numberRound %in% 1)
secondRound <- subset(draft, numberRound %in% 2)

draft <- subset(draft, select = -numberRound)

firstRoundStats <- left_join(firstRound, stats, by = c("namePlayer" = "Player"))
secondRoundStats <- left_join(secondRound, stats, by = c("namePlayer" = "Player"))

secondRoundStats <- subset(secondRoundStats, secondRoundStats$count <= 3)

stats <- as_tibble(rbind(firstRoundStats, secondRoundStats))

```
Now, we can get the Sum VORP for each player for the duration of their rookie
contract.  

```{r echo=TRUE}
sumVORP <- stats %>% 
  group_by(namePlayer) %>% 
  summarise(sumPlayerVORP = sum(VORP)) %>%
  mutate_if(is.numeric, round, 3)

draft <- left_join(draft, sumVORP, by = "namePlayer")
```
For every player didn't end up playing a game in the league, we will assign them a value of replacement-level value of 0.

```{r echo=TRUE}
draft[is.na(draft)] <- 0
```
Now that we have the relevant data joined into one table, let's look at 
that table.

```{r echo=TRUE}
head(draft)
```
In looking at the data, Isaiah Thomas being an impactful NBA player taken with the 60th pick is such an outlier that it skews the pick valuations later on.  Because of this, I will remove him from the dataset to provide a more accurate reflection of the value of late second round picks.

```{r echo=TRUE}
draft <- subset(draft, draft$namePlayer != 'Isaiah Thomas')
```

## Analysis

Now that we have the Summed VORP for each player, we can get the Mean VORP for
each pick.

```{r echo=TRUE}
pickScore <- draft %>% 
  group_by(numberPickOverall) %>% 
  summarise(meanPickVORP = mean(sumPlayerVORP)) %>%
  mutate_if(is.numeric, round, 3)

ggplot(pickScore, aes(numberPickOverall, meanPickVORP)) +
  geom_point() +
  theme_minimal() +
  labs(title = 'Mean Value Over Replacement Player For Each Pick', 
              x='Pick Number', y='Mean VORP') + geom_smooth(method = 'scam',
                                                            formula = y ~ s(x, k = 5,
                                                            bs = "mdcx"), se=TRUE)

```

To account for outliers, I performed monotonic smoothing to fit a curve through the
data points.  It gives us the values from the blue curve in the above graph.

```{r echo=TRUE}
pickScore$pickValue <-predict(scam(pickScore$meanPickVORP~s(pickScore$numberPickOverall,k=5,bs="mdcx")),se=FALSE)

pickScore <- pickScore[, -c(2)]

```
To make the pick values more practical and applicable for our purposes,
I normalized the values so that the first pick was worth 100 points.

```{r echo=TRUE}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

pickScore$pickValue <- normalize(pickScore$pickValue)
pickScore$pickValue <- round(pickScore$pickValue*100,1)
```
To see the calculated values of each pick, please see Table 1 below.

```{r echo=TRUE}
round1 <- pickScore[1:30,]
round2 <- pickScore[31:60,]

kable(list(round1, round2), 
             caption = "NBA Draft Pick Value Table",
             col.names = c("Pick Number", "Pick Value"),
             format = 'html', booktabs = TRUE) %>%
  kable_styling(bootstrap_options = 'striped', full_width = FALSE, position='left')

```
Below please find the final graph.

```{r echo=TRUE}
ggplot(pickScore, aes(numberPickOverall, pickValue)) +
       geom_point() +
       theme_minimal() +
       labs(title = 'NBA Draft Pick Value Chart', 
       x='Pick Number', y='Pick Value')

```

The values in the above curve make intuitive sense, as we see a steep dropoff in value throughout the lottery, continuing through to about the 20th pick. The curve then steadily declines through the rest of the draft, with the slope declining early into the second round.

## Conclusion

I think this research can be of major importance to NBA decision makers.  Having quantitative data on what each pick has historically produced makes it easier to understand the draft.  The values of each pick can be used in trade negotiation talks to see if historical data indicates whether a team should or shouldn't make a deal.  That being said, understanding the general curve is more important than the exact values.  

To use a real world example, let us look at the Orlando Magic, who have the 5th and 8th pick in this year's draft.  The summed value of those two picks is 125.6.  There are rumors that they are interested in trading those two picks for the second overall pick, which is worth 92.1 points.  

From a pure pick value standpoint, the Magic would be very much overpaying in this deal.  However, with all of these potential trades, it of course depends on the Magic's valuation of who they would be taking at 2 versus who they would be taking at 5 and 8. If we were to look at the New York Knicks, who have the 19th and 21st picks in the draft, the sum of those two picks is about equivalent to the 12th pick.

## Future Research 

One thing to consider for future research is to see how certain player archetypes tend to perform based on where they were selected. Additionally, the scope of this study would be greatly enhanced with a better understanding of the value of future draft picks relative to current draft picks. As many teams across the league look to stockpile future draft capital, this would be extremely valuable information in guiding each team's strategy.  This would require sound predictive modelling of how good each team will be x years in advance.

Without knowing exact pick numbers years in advance, it might be beneficial to look at pick ranges instead of pick numbers. Whereas teams won’t know exactly where they will pick, they would likely have an educated understanding of where they will be picking in future years, especially for presumptive playoff teams. There are a lot of different avenues that this research can take us with regards to deciding upon optimal basketball strategy.